Real-space computation of $E$/$B$-mode maps I: Formalism and Compact Kernels

### 2017-11-23 articles/formalism/sections/real_space_formalism.tex
Sec 3.1 is now more more complete and satisfactory at some level. Still needs refining, but the core ideas are now in place.

where the symbol ${}_{0}Y_e$ is used to denote the sub-matrix ${}_{0}Y_{\hat{n}_e \times \ell m} \equiv {}_{0}Y_{\ell m}(\hat{n}_e)$, the symbol ${}_{\pm 2}Y^{T*}_q$ is used to denote the matrix ${}_{\pm 2}Y^*_{\ell m \times \hat{n}_q} \equiv {}_{\pm 2}Y^*_{\ell m}(\hat{n}_q)$ and the summation is over the multipole indices $\ell,m$. \revisit{We have purposefully used the index ``e'' to denote the location where the scalar fields are measured and the index ``q'' to denote the location where the Stokes parameters are measured.} Using the conjugation properties of the spin spherical harmonic functions it can be shown that the following relation holds true,

@@

\textit{Convolution:} Working from the frame of the pixel ``e'' where the scalar modes need to be evaluated, the Euler angles to the surrounding pixels are: $\alpha_{eq}, \beta_{eq},\gamma_{eq}$. In terms of these Euler angles, the value of scalar fields $E$ \& $B$ at the central pixel ``e" can be evaluated by convolving over the measured Stokes $Q$ \& $U$ parameters in the surrounding pixels marked by indices ``q''. Specifically, the convolution is fully defined in terms of the Euler angles $\beta_{eq}~ \& ~\gamma_{eq}$ and is given by the following equation,
%
\beq \label{eq:qu2eb_convolution_explicit}
\bmat E_e \\ B_e  \emat =- \sum_{q=1}^{N_{\rm pix}}{{}_{\mm}f}(\beta_{eq},\ell_{\rm min},\ell_{\rm max})\bmat \cos(2 \gamma_{eq}) & -\sin(2\gamma_{eq})\\  \sin(2 \gamma_{eq})  & \cos(2 \gamma_{eq}) \emat  \bmat Q_q \\ U_q  \emat \Delta \Omega\,,
\eeq
%
where $\Delta \Omega$ denotes the pixel area and all the symbols have their usual meaning. 
%\revisit{This has an elegant interpretation: to derive the E and/or B field at any given position we need to find the cosine quadrupole transform and the sine quadrupole transform of the Stokes Q \& U parameters on circles around this position, weigh the transform by the value of the function $f(\beta,\ell_{\rm min},\ell_{\rm max})$, $\beta$ being the radius of the circle and sum up the results with appropriate signs, to construct the respective scalar fields.} 
When written in terms of the complex spin-0 field $[E + iB]$ and the spin-2 field ${}_{+2}X$, the above equation can be expressed more concisely as follows,
%
\begin{subequations} \label{q:qu2eb_convolution_concise}
\beqry 
%[E + iB](\hat{n}_e) &=& - \Delta \Omega \sum_{q=1}^{N_{\rm pix}} \left(\sum_{\ell=\ell_{\rm min}}^{\ell_{\rm max}} \frac{2 \ell+1}{4 \pi} \sqrt{\frac{(\ell-2)!}{(\ell+2)!}} P_{\ell}^{2}(\beta_{eq}) \right) {\Bigg( e^{i2 \gamma_{eq}}   {}_{+2}X (\hat{n}_q) \Bigg)} \,, \label{eq:qu2eb_physical}\\
[E + iB](\hat{n}_e) &=& \Bigg\lbrace \left[ - \Delta \Omega \sum_{\ell=\ell_{\rm min}}^{\ell_{\rm max}} \sqrt{\frac{2 \ell+1}{4 \pi}}Y_{\ell 2}(\beta_{eq},\gamma_{eq}) \right]  \circ {}_{+2}X \Bigg\rbrace(\hat{n}_e) \,, \\
&=& \Bigg\lbrace \mathcal{M}_{Beam} \circ {}_{+2}X \Bigg\rbrace(\hat{n}_e)\,, \label{eq:qu2eb_convolution} 
\eeqry
\end{subequations}
%
where $\circ$ denotes the convolution and the spherical harmonic functions denote the rotated functions such that the pole of the function coincides with the direction $\hat{n}_e$. %and the reference zero for the azimuthal angle is the local longitude $\phi_0$.

\textit{Radiating:} On the contrary, while working from the frame of the pixel ``q'', the Euler angles to the surrounding pixels are: $\alpha_{qe}, \beta_{qe},\gamma_{qe}$. The E \& B mode maps resulting from the Stokes Q \& U parameters specified at location $\hat{n}_q$ is given by the following expression,
%
\beq  \label{eq:qu2eb_radiation_explicit}
\bmat E_e \\ B_e  \emat_{q} =- {{}_{\mm}f}(\beta_{qe},\ell_{\rm min},\ell_{\rm max})\bmat \cos(2 \alpha_{q e}) & \sin(2\alpha_{q e})\\  -\sin(2 \alpha_{q e})  & \cos(2 \alpha_{q e}) \emat  \bmat Q_{q} \\ U_{q}  \emat \Delta \Omega \,.
\eeq
%
The total map of E \& B modes can be simply evaluated by summing over the contribution from the Stokes parameters at each location $\hat{n}_q$ : $\bar{S} = \sum_{q=1}^{N_{\rm pix}} \bar{S}_q$. The above expression can again be cast in a more concise form as follows,
%
\begin{subequations} \label{eq:qu2eb_radiation_concise}
\beqry 
\left[E + iB\right] &=& -\Delta \Omega  \sum_{q=1}^{N_{\rm pix}} \Big[ {}_{+2}X(\hat{n}_{q}) e^{-i2\alpha_{q e}} \Big]  {{}_{\mm}f}(\beta_{q e}) \,, \\
&=&    \sum_{q=1}^{N_{\rm pix}} \Bigg\lbrace {}_{+2}X(\hat{n}_{q}) \cdot \left[ - \Delta \Omega \sum_{\ell=\ell_{\rm min}}^{\ell_{\rm max}} \sqrt{\frac{2 \ell+1}{4 \pi}}Y^*_{\ell 2}(\beta_{qe},\alpha_{qe}) \right] \Bigg\rbrace \,, \\
&=& \sum_{q=1}^{N_{\rm pix}} {}_{+2}X(\hat{n}_{q}) \cdot  \mathcal{M}_{Green}(\hat{n}_{q}) \,,
\eeqry
\end{subequations}
%
where ``$\cdot$'' denotes a simple scalar multiplication. \revisit{$\mathcal{M}_{Green}(\hat{n}_{q})$ is response of the real space operator $\bar{O}$ when operated on a Stokes field: $\delta(\hat{n}-\hat{n}_q) + i \delta(\hat{n}-\hat{n}_q)$, it can be thought of as the Green's function of the real space operator.} \comment{Is this interpretation correct ? Is the complex delta function a spin 2 function ?}

The convolution kernel has a azimuthal part which depends only on the Euler angle $\alpha$ and has no multipole dependence and this is the primary operation which translates between the two different spin representation of CMB polarization. The radial part of the kernel is specified by the function $f(\beta,\ell_{\rm min},\ell_{\rm max})$, it depends only on the angular separation given by the Euler angle $\beta$ and it completely incorporates the multipole $\ell$ dependence of the kernel. It is the radial part of the kernel that completely determines the locality of this operator. Note that this kernel does not depend on the Euler angle $\gamma$ \comment{Why is that and how do you understand this ? }.

We know that the product of two functions with spins $s_1$ and $s_2$ results in a function with spin $s_1 + s_2$: ${}_{s_1 +s_2}f = {}_{s_1}g \,{}_{s_2}h$. Since $[Q + iU]$ is a field with spin +2 and the field $\exp(-i2\alpha)$ has spin -2, the resultant field formed by the product of these two function has spin-0. This makes intuitive, the construction of the spin-0 E and B modes of polarization. Note that the coordinate dependence of the Stokes parameters cannot be integrated out at the location $\hat{n}_0$ where the scalar fields are to be evaluated and at the diametrically opposite location, since the azimuthal angle $\alpha$ becomes ill-defined at these points. The convolution kernel has vanishing contribution from these points, since the radial part of the kernel is given by a weighted linear combination of $P_{\ell}^2$ Legendre polynomials which $P_{\ell}^2(\beta) \propto \sin^2{\beta} \rightarrow 0 $ as $\beta \rightarrow 0 ,\pi$.  \revisit{It is also interesting to note that the E-modes are constructed by product of functions $(U\sin{2 \alpha}, Q\cos{2\alpha})$ which have the same parity and hence have even parity while the B-modes are constructed by multiplying functions $(Q\sin{2 \alpha}, U\cos{2\alpha})$ of opposite parity and hence have an odd parity.}

>>>>

where the symbol ${}_{0}Y_e$ is used to denote the sub-matrix ${}_{0}Y_{\hat{n}_e \times \ell m} \equiv {}_{0}Y_{\ell m}(\hat{n}_e)$, the symbol ${}_{\pm 2}Y^{T*}_q$ is used to denote the matrix ${}_{\pm 2}Y^*_{\ell m \times \hat{n}_q} \equiv {}_{\pm 2}Y^*_{\ell m}(\hat{n}_q)$ and the summation is over the multipole indices $\ell,m$. \revisit{We purposefully use the index ``e'' to denote the location where the scalar fields are measured and the index ``q'' to denote the location where the Stokes parameters are measured.} Using the conjugation properties of the spin spherical harmonic functions it can be shown that the following relation holds true,

@@

\textit{Convolution:} Working from the frame of the pixel ``e'' where the scalar modes need to be evaluated, the Euler angles to the surrounding pixels are: $\alpha_{eq}, \beta_{eq},\gamma_{eq}$. In terms of these Euler angles, the value of scalar fields $E$ \& $B$ at the central pixel ``e" can be evaluated by convolving over the measured Stokes $Q$ \& $U$ parameters in the surrounding pixels marked by indices ``q''. Specifically, the convolution is fully defined in terms of the Euler angles $\beta_{eq}~ \& ~\gamma_{eq}$ and is given by the following equation,
%
\beq \label{eq:qu2eb_convolution_explicit}
\bmat E_e \\ B_e  \emat =- \sum_{q=1}^{N_{\rm pix}}{{}_{\mm}f}(\beta_{eq},\ell_{\rm min},\ell_{\rm max})\bmat \cos(2 \gamma_{eq}) & -\sin(2\gamma_{eq})\\  \sin(2 \gamma_{eq})  & \cos(2 \gamma_{eq}) \emat  \bmat Q_q \\ U_q  \emat \Delta \Omega\,,
\eeq
%
where $\Delta \Omega$ denotes the pixel area and all the symbols have their usual meaning. 
%\revisit{This has an elegant interpretation: to derive the E and/or B field at any given position we need to find the cosine quadrupole transform and the sine quadrupole transform of the Stokes Q \& U parameters on circles around this position, weigh the transform by the value of the function $f(\beta,\ell_{\rm min},\ell_{\rm max})$, $\beta$ being the radius of the circle and sum up the results with appropriate signs, to construct the respective scalar fields.} 
When written in terms of the complex spin-0 field $[E + iB]$ and the spin-2 field ${}_{+2}X$, the above equation can be expressed more concisely as follows,
%
\begin{subequations} \label{q:qu2eb_convolution_concise}
\beqry 
[E + iB](\hat{n}_e) &=& - \Delta \Omega \sum_{q=1}^{N_{\rm pix}} \left(\sum_{\ell=\ell_{\rm min}}^{\ell_{\rm max}} \frac{2 \ell+1}{4 \pi} \sqrt{\frac{(\ell-2)!}{(\ell+2)!}} P_{\ell}^{2}(\beta_{eq}) \right) {\Bigg( e^{i2 \gamma_{eq}}   {}_{+2}X (\hat{n}_q) \Bigg)} \,, \label{eq:qu2eb_physical}\\
&=& \Bigg\lbrace \left[ - \Delta \Omega \sum_{\ell=\ell_{\rm min}}^{\ell_{\rm max}} \sqrt{\frac{2 \ell+1}{4 \pi}}Y_{\ell 2}(\beta_{eq},\gamma_{eq}) \right]  \circ {}_{+2}X \Bigg\rbrace(\hat{n}_e) \,, \\
&=& \Bigg\lbrace \mathcal{M}_{Beam} \circ {}_{+2}X \Bigg\rbrace(\hat{n}_e)\,, \label{eq:qu2eb_convolution} 
\eeqry
\end{subequations}
%
where $\circ$ denotes the convolution and the spherical harmonic functions denote the rotated functions such that the pole of the function coincides with the direction $\hat{n}_e$. %and the reference zero for the azimuthal angle is the local longitude $\phi_0$.

\textit{Radiating:} On the contrary, while working from the frame of the pixel ``q'', the Euler angles to the surrounding pixels are: $\alpha_{qe}, \beta_{qe},\gamma_{qe}$. The E \& B mode maps resulting from the Stokes Q \& U parameters specified at location $\hat{n}_q$ is given by the following expression,
%
\beq  \label{eq:qu2eb_radiation_explicit}
\bmat E_e \\ B_e  \emat_{q} =- {{}_{\mm}f}(\beta_{qe},\ell_{\rm min},\ell_{\rm max})\bmat \cos(2 \alpha_{q e}) & \sin(2\alpha_{q e})\\  -\sin(2 \alpha_{q e})  & \cos(2 \alpha_{q e}) \emat  \bmat Q_{q} \\ U_{q}  \emat \Delta \Omega \,.
\eeq
%
The total map of E \& B modes can be simply evaluated by summing over the contribution from the Stokes parameters at each location $\hat{n}_q$ : $\bar{S} = \sum_{q=1}^{N_{\rm pix}} \bar{S}_q$. The above expression can again be cast in a more concise form as follows,
%
\begin{subequations} \label{eq:qu2eb_radiation_concise}
\beqry 
\left[E + iB\right] &=& -\Delta \Omega  \sum_{q=1}^{N_{\rm pix}} \Big[ {}_{+2}X(\hat{n}_{q}) e^{-i2\alpha_{q e}} \Big]  {{}_{\mm}f}(\beta_{q e}) \,, \\
&=&    \sum_{q=1}^{N_{\rm pix}} \Bigg\lbrace {}_{+2}X(\hat{n}_{q}) \cdot \left[ - \Delta \Omega \sum_{\ell=\ell_{\rm min}}^{\ell_{\rm max}} \sqrt{\frac{2 \ell+1}{4 \pi}}Y^*_{\ell 2}(\beta_{qe},\alpha_{qe}) \right] \Bigg\rbrace \,, \\
&=& \sum_{q=1}^{N_{\rm pix}} {}_{+2}X(\hat{n}_{q}) \cdot  \mathcal{M}_{Green}(\hat{n}_{q}) \,,
\eeqry
\end{subequations}
%
where ``$\cdot$'' denotes a simple scalar multiplication. \revisit{$\mathcal{M}_{Green}(\hat{n}_{q})$ is response of the real space operator $\bar{O}$ when operated on a Stokes field: $\delta(\hat{n}-\hat{n}_q) + i \delta(\hat{n}-\hat{n}_q)$, hence it can be thought of as the Green's function of the real space operator.} \comment{Is this interpretation correct ? Is the complex delta function a spin 2 function ?}

Recall that the product of two functions with spins $s_1$ and $s_2$ results in a function with spin $s_1 + s_2$: ${}_{s_1 +s_2}f = {}_{s_1}g \,{}_{s_2}h$ \comment{The following argument is that of the product of a number with a field, so is this relevant ? }. If one rotates by an angle $\phi$ about the local $\hat{z}_q$ axis, the local Stokes parameters transform as: ${}_{+2}X(\hat{n}_q) \xrightarrow{\mathcal{R}_{\hat{z}_q}(\phi)} {}_{+2}X(\hat{n}_q) e^{i2\phi} $. Since this angle $\phi$ can be thought of as the phase angle $\gamma$ in the $z-y_1-z_2$ convention of Euler angles, one can see that: $e^{-i2\alpha_{qe}} \xrightarrow{\mathcal{R}_{\hat{z}_q}(\phi)} e^{-i2\alpha_{qe}} e^{-i2\phi}$. The Euler angles $|\beta_{qe}|$ which measure the angular distance between pixels remains unaltered under this local rotation operation.  Given these transformation properties one can now note that the product ${}_{+2}X(\hat{n}_q)e^{-i2\alpha_{qe}}$ is invariant under rotations  $\mathcal{R}_{\hat{z}_q}(\phi)$ and hence must form a spin-0 field.  This argument makes intuitive, the construction of the spin-0 E and B modes of polarization from the measured Stokes parameters. \revisit{Here it is also interesting to note that the E-modes are constructed by product of functions $(U\sin{2 \alpha}, Q\cos{2\alpha})$ which have the same parity and hence have even parity while the B-modes are constructed by multiplying functions $(Q\sin{2 \alpha}, U\cos{2\alpha})$ of opposite parity and hence have an odd parity.}

Using \eq{eq:qu2eb_radiation_explicit} one can see that when $q=e$: $E_q=-{{}_{\mm}f}(\beta_{qq})Q_q$ and $B_q = -{{}_{\mm}f}(\beta_{qq})U_q$ which does not transform as a spin-0 field under local rotations unless ${{}_{\mm}f}(\beta_{qq}=0)=0$. One can make a similar argument when pixels ``q'' and ``e'' are diametrically opposite. Therefore the radial kernels have to satisfy the constraint that they vanish at $\beta = 0, \pi$. The radial kernel ${{}_{\mm}f}(\beta)$ is constructed by forming a weighted linear combination of $P_{\ell}^2$ Legendre polynomials which have the following limiting form: $P_{\ell}^2(\beta) \propto \sin^2{\beta} \rightarrow 0 $ as $\beta \rightarrow 0 ,\pi$ and therefore satisfy this necessary constraint.

The real space operator has a azimuthal part which depends only on the Euler angle $\alpha_{qe} (\gamma_{eq})$, has no multipole $\ell$ dependence and is the crucial operation which translates between the two different spin representation of CMB polarization. The radial part of the kernel is specified by the function ${}_{\mm}f(\beta,\ell_{\rm min},\ell_{\rm max})$, it depends only on the angular separation $|\beta_{eq}|$ between pixels and completely incorporates the multipole $\ell$ dependence of the kernel. It's the radial part of the kernel that determines the locality of the real space operator.
